Isomorphisms as Groupoids: \\ Homotopy Type Theory and Fractional Types

### 2013-11-07 f.tex
new versions of agda code compatible with 3.3

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{url}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage[sans]{dsfont}
\usepackage{color, colortbl}
\usepackage{courier}
\usepackage{graphicx}
\usepackage[all]{xy}
\xyoption{line}

@@

\newcommand{\br}{\mathit{Br}}
\newcommand{\ssep}{\hspace{1em}} 
\newcommand{\sym}{c}       % braid: could be \sigma or c or \gamma
\newcommand{\cp}{\circ}                 % composition
\newcommand{\symi}{\sym\inv}
\newcommand{\inv}{^{-1}}
\newcommand{\lolli}{\multimap} 
\newcommand{\id}{\textrm{\rm id}}       % identity
\newcommand{\x}{\otimes}
\newcommand{\boolt}{\mathit{bool}}
\newcommand{\ffv}{\mathit{false}}
\newcommand{\ttv}{\mathit{true}}
\newcommand{\notb}{\mathit{not}}
\newcommand{\union}{\cup}
\newcommand{\transport}[2]{\overline{#1}(#2)}
\newcommand{\refl}{\mathit{refl}}
\renewcommand{\path}{\leadsto}
\newcommand{\paths}[1]{\mathds{P}(#1)}
%% \renewcommand{\circ}{\circlearrowleft}
\newcommand{\alt}{~|~}
\newcommand{\dt}[1]{\llbracket #1 \rrbracket}
\newcommand{\leftv}[1]{\textsf{left}~#1}
\newcommand{\rightv}[1]{\textsf{right}~#1}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\identlp}{\mathit{identl}_+}
\newcommand{\identrp}{\mathit{identr}_+}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{identl}_*}
\newcommand{\identrt}{\mathit{identr}_*}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\distz}{\mathit{dist}_0}
\newcommand{\factorz}{\mathit{factor}_0}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\idc}{\mathit{id}}
\newcommand{\wedgesum}{\vee}
\newcommand{\smashproduct}{\wedge}
\newcommand{\symc}[1]{\mathit{sym}~#1}
\newcommand{\proves}{\vdash}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}

@@

>>>>

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{tikz}
\usepackage{url}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage[sans]{dsfont}
\usepackage{color, colortbl}
\usepackage{courier}
\usepackage{graphicx}
\usepackage[all]{xy}
\xyoption{line}

@@

\newcommand{\br}{\mathit{Br}}
\newcommand{\ssep}{\hspace{1em}} 
\newcommand{\sym}{c}       % braid: could be \sigma or c or \gamma
\newcommand{\cp}{\circ}                 % composition
\newcommand{\pc}{\fatsemi}                 % path composition
\newcommand{\symi}{\sym\inv}
\newcommand{\inv}{^{-1}}
\newcommand{\lolli}{\multimap} 
\newcommand{\id}{\textrm{\rm id}}       % identity
\newcommand{\x}{\otimes}
\newcommand{\boolt}{\mathit{bool}}
\newcommand{\ffv}{\mathit{false}}
\newcommand{\ttv}{\mathit{true}}
\newcommand{\notb}{\mathit{not}}
\newcommand{\union}{\cup}
\newcommand{\transport}[2]{\overline{#1}(#2)}
\newcommand{\refl}{\mathit{refl}}
\renewcommand{\path}{\leadsto}
\newcommand{\paths}[1]{\mathds{P}(#1)}
%% \renewcommand{\circ}{\circlearrowleft}
\newcommand{\alt}{~|~}
\newcommand{\dt}[1]{\llbracket #1 \rrbracket}
\newcommand{\leftv}[1]{\textsf{left}~#1}
\newcommand{\rightv}[1]{\textsf{right}~#1}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\identlp}{\mathit{identl}_+}
\newcommand{\identrp}{\mathit{identr}_+}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{identl}_*}
\newcommand{\identrt}{\mathit{identr}_*}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\distz}{\mathit{dist}_0}
\newcommand{\factorz}{\mathit{factor}_0}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\idc}{\mathit{id}}
\newcommand{\wedgesum}{\vee}
\newcommand{\smashproduct}{\wedge}
\newcommand{\symc}[1]{\mathit{sym}~#1}
\newcommand{\proves}{\vdash}
\newcommand{\ap}[2]{\mathit{ap}~#1~#2}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}
\newcommand{\adjoint}[1]{#1^{\dagger}}

@@

Univalence says:
\[ 
(A \simeq B) \quad\simeq\quad (A \equiv B)
\]
where $\simeq$ is a strong form of isomorphism witnessed by functions from
$A$ to $B$ and vice versa that are inverses and $\equiv$ is the propositional
equality in type theoretic terms or a path (homotopy) in topological
terms. The axiom relates two notions that are a priori different: functions
and paths. However if we define functions from the outset as paths then there
is no need for this axiom.

So what would it take to have each function correspond to a path and
higher-order functions correspond to paths between paths etc.? Since paths
form an equivalence relation, the first immediate observation is that all
functions must be reversible. Formally we need the following properties of
paths to be satisfied by functions. (We write $\path$ for paths instead of
$\equiv$.)

\begin{itemize}
\item For each $A$, there is function $\id : A \path A$,

\item for each function $f: A \path B$, there exists a function $\adjoint{f}
  : B \path A$, and

\item for each functions $f : A \path B$ and $g : B \path C$, there exists a
  function $g \cp f : A \path C$, such that:

\begin{itemize}
\item $f \path f \cp \id$;
\item $f \path \id \cp f$;
\item $\adjoint{f} \cp f \path \id$;
\item $f \cp \, \adjoint{f} \path \id$;
\item $\adjoint{(\adjoint{f})} \path f$;
\item $f \cp (g \cp h) \path (f \cp g) \cp h$
\end{itemize}
\end{itemize}
Similar equivalences hold are higher-levels. A good starting point for this
notion of functions are the $\Pi$-combinators we previously developed. Each
combinator in that setting had an \emph{adjoint} going in the other direction
that acts as the inverse. Moreover if $f : A \path B$ then $A$ must be
isomorphic to $B$ and $f$ is permutation. We use $!p$ for the inverse of a
path.

But first, let's examine the structure of a path $f : A \path B$. A natural
representation for a path representing a function is a collection of paths
(called \emph{fibers}) one for each $a : A$. So for example, if $A$ is the
set $\{ x , y , z \}$ and $B$ is the set $\{ a , b , c \}$ (remember that $A$
and $B$ must be isomorphic), and $f$ is the function mapping $x$ to $a$, $y$
to $b$, and $z$ to $c$, then $f : A \path B$ is the collection of paths $f_x
: x \path a$, $f_y : y \path b$, and $f_z : z \path c$.

Assuming this representation of functions, let's see what properties are
induced for these functions.

The first property is that functions should preserve paths, i.e., if there is
a path between $x$ and~$y$ then there should be a path between $f(x)$
and~$f(y)$. From the figure below it is clear that:
\[ 
\ap{f}{p} = \, !f_x \pc p \pc f_y
\]

\begin{center}
\begin{tikzpicture}
  \draw (-3,0) ellipse (1.5cm and 3cm);
  \draw (3,0) ellipse (1.5cm and 3cm);
  \draw[fill] (-3,1.5) circle [radius=0.025];
  \draw[fill] (-3,-1.5) circle [radius=0.025];
  \node[above] at (-3,1.5) {$x$};
  \node[below] at (-3,-1.5) {$y$};
  \draw[fill] (3,1.5) circle [radius=0.025];
  \draw[fill] (3,-1.5) circle [radius=0.025];
  \node[above] at (3,1.5) {$f(x)$};
  \node[below] at (3,-1.5) {$f(y)$};
  \draw[->,cyan,thick] (-3,1.5) -- (-3,-1.5);
  \node[left,cyan] at (-3,0) {$p$};
  \draw[->,dashed,cyan,thick] (3,1.5) -- (3,-1.5);
  \node[right,cyan] at (3,0) {$\mathit{ap}~f~p$};
  \draw[->,red,thick] (-3,1.5) -- (3,1.5);
  \draw[->,red,thick] (-3,-1.5) -- (3,-1.5);
  \node[red,below] at (0,1.5) {$f_x$};
  \node[red,below] at (0,-1.5) {$f_y$};
\end{tikzpicture}
\end{center}

Let $p : x \path y$ and $q : y \path z$, then the following properties 
hold:
\begin{itemize}
\item $\ap{f}{(p \pc q)} \path \ap{f}{p} \pc \ap{f}{q}$:
\[\begin{array}{rcl}
LHS &=& \, !f_x \pc (p \pc q) \pc f_z \\
    &\path& ! f_x \pc p \pc f_y \pc \, ! f_y \pc q \pc f_z \\
    &\path& (! f_x \pc p \pc f_y) \pc (! f_y \pc q \pc f_z) \\
    &=& RHS
\end{array}\]
\item $\ap{f}{!p} \path \, !(\ap{f}{p})$
\[\begin{array}{rcl}
LHS &=& \, !f_y \pc (!p) \pc f_x \\
    &\path& \, !(!f_x \pc p \pc f_y) \\
    &=& RHS
\end{array}\]
\item $\ap{g}{(\ap{f}{p})} \path \ap{(g \cp f)}{p}$
\[\begin{array}{rcl}
LHS &=& \, !g_{f_x} \pc !f_x \pc p \pc f_y \pc g_{f_y} \\
    &\path& \, ! (f_x \pc g_{f_x}) \pc p \pc (f_y \pc g_{f_y}) \\
    &=& RHS
\end{array}\]
\item $\ap{\id}{p} \path p$
\[\begin{array}{rcl}
LHS &=& \, !\id_x \pc p \pc \id_y \\
    &\path& p \\
    &=& RHS
\end{array}\]
\end{itemize}

A homotopy between functions $f$ and $g$ is a 2-path between each pair of
fibers $f_x$ and $g_x$. These 2-paths are induced by the following diagram:

\begin{center}
\begin{tikzpicture}
  \draw[fill] (-3,0) circle [radius=0.025];
  \node[below] at (-3,0) {$x$};
  \draw[fill] (3,1.5) circle [radius=0.025];
  \node[above] at (3,1.5) {$a$};
  \draw[fill] (3,-1.5) circle [radius=0.025];
  \node[below] at (3,-1.5) {$b$};
  \draw[->,cyan,thick] (-3,0) -- (3,1.5);
  \draw[->,cyan,thick] (-3,0) -- (3,-1.5);
  \draw[->,red,thick] (3,1.5) -- (3,-1.5);
  \node[red,above] at (0,0.5) {$f_x$};
  \node[red,below] at (0,-0.5) {$f_y$};
  \draw[->,double,dashed,blue,thick] (1.5,1) to [out=225, in=135] (1.5,-1);
  \node[blue,right] at (1.5,0) {$H_x$};
\end{tikzpicture}
\end{center}

Here is something more interesting. Let $p : y \path z$ in $A$. Let $f$ and
$g$ be functions from $A$ to $B$. Let $H$ be a 2-path between $f_y$ and
$g_y$. If we are given a path $f(y) \path g(y)$ in B, then we can construct a
2-path between $f_z$ and $g_z$. (This is Thm 2.11.3 in the book.) We do this
going backwards along $f_z$, backwards along $p$, then follow $f_y$ and the
path to $g(y)$, and backwards along $g_y$, forwards along $p$, and finally
forwards along $g_z$:

\begin{center}
\begin{tikzpicture}
  \draw[fill] (-3,0) circle [radius=0.025];
  \node[below] at (-3,0) {$x$};
  \draw[fill] (3,1.5) circle [radius=0.025];
  \node[above] at (3,1.5) {$a$};
  \draw[fill] (3,-1.5) circle [radius=0.025];
  \node[below] at (3,-1.5) {$b$};
  \draw[->,cyan,thick] (-3,0) -- (3,1.5);
  \draw[->,cyan,thick] (-3,0) -- (3,-1.5);
  \draw[->,red,thick] (3,1.5) -- (3,-1.5);
  \node[red,above] at (0,0.5) {$f_x$};
  \node[red,below] at (0,-0.5) {$f_y$};
  \draw[->,double,dashed,blue,thick] (1.5,1) to [out=225, in=135] (1.5,-1);
  \node[blue,right] at (1.5,0) {$H_x$};
  \draw[fill] (-3,-5) circle [radius=0.025];
  \node[below] at (-3,-5) {$x$};
  \draw[fill] (3,-3.5) circle [radius=0.025];
  \node[above] at (3,-3.5) {$a$};
  \draw[fill] (3,-6.5) circle [radius=0.025];
  \node[below] at (3,-6.5) {$b$};
  \draw[->,cyan,thick] (-3,-5) -- (3,-3.5);
  \draw[->,cyan,thick] (-3,-5) -- (3,-6.5);
  \draw[->,red,thick] (3,-3.5) -- (3,-6.5);
  \node[red,above] at (0,-4.5) {$f_x$};
  \node[red,below] at (0,-5.5) {$f_y$};
\end{tikzpicture}
\end{center}

Now the most interesting question is whether we can internalize functions as
objects. Given $f$ a function from $A$ to $B$, what we can do below is to
represent each fiber $f_x$ as an element of the type $(\frac{1}{A} \times
B)$. As an example, let $A$ be the set $\{x,y,z\}$ and let $B$ be
$\{a,b,c\}$, the type $\frac{1}{A}$ has one element $\frac{1}{3}$ which is
one point that has three self loops labeled $x \path$, $y \path$, and $z
\path$. A value $(\frac{1}{3},a)$ is a fiber of the function that produces
$a$ as the second component indicates. The first component states that the
$a$ could have come from $x$ along the path $x \path$, or from $y$ along the
path $y \path$ or from $z$ along the path $z \path$.

\newpage
